---
title: "Yavorsky, Honka & Chen (2021) vs SIFT"
subtitle: "What does normalizing the post-search spread to one actually cost?"
format: html
bibliography: ../references.bib
execute: {echo: false, warning: false}
---

```{r}
#| include: false
res <- function(n) {
  p <- file.path("out", paste0(n, ".rds"))
  if (!file.exists(p)) p <- file.path("comparisons", "out", paste0(n, ".rds"))
  readRDS(p)
}
```

## The question

@Yavorsky_2021 show that the match-value standard deviation — the post-search
spread, $\tilde\sigma$ — can be estimated rather than normalized to one, given an
exogenous search-cost shifter, and that the normalization does real damage: they
estimate 8.16 against the conventional value of one, and imposing one reverses
the sign of a search-cost coefficient and understates a counterfactual by a
factor of three to four.

They have the shifter but no panel. @Morozov_2021 have the panel but normalize
the spread. SIFT has both, plus a designed product page that fixes part of the
spread rather than inferring it.

Two things to establish. **What does normalizing cost**, as the true spread moves
away from one? And **how much of the benefit comes from the panel** as opposed to
the shifter?

## Design

Sweep the true spread across $\tilde\sigma \in \{1.29, 2.58, 5.15\}$, spanning
the conventional normalization and the neighbourhood of Yavorsky's field
estimate. At each level fit the model twice: once with $\tilde\sigma$ free, once
with it forced to 1.

"Normalizing" is implemented the way a researcher actually does it —
$\tilde\sigma$ is imposed in the **reservation value**, which is the search rule,
while $\boldsymbol\kappa$ and $\sigma_\xi$ continue to govern realized utility.

Total tasks are held fixed at 400 across the $T$ arms, so $T=1$ against $T=8$
isolates the panel rather than confounding it with sample size. $T=1$ is the
closest simulated analogue of Yavorsky's design.

The counterfactual mirrors their at-home-test-drive exercise: halve search costs
and predict the change in clicks. Code: [`exp_mvsd.R`](exp_mvsd.R).

## Result 1 — the spread is recovered when estimated

```{r}
r <- res("mvsd")$res
knitr::kable(data.frame(
  `true sigma`  = sprintf("%.2f", r$sig_true),
  `T`           = r$T,
  `sigma hat`   = sprintf("%.2f", r$sig_hat_free),
  `gamma0 true` = sprintf("%.2f", r$gamma0_true),
  `gamma0 free` = sprintf("%.2f", r$gamma0_free),
  `gamma0 normalized` = sprintf("%.2f", r$gamma0_norm),
  check.names = FALSE), align = "r")
```

With $\tilde\sigma$ free, it is recovered across the whole range — 1.24–1.61
against a truth of 1.29, 2.30–2.59 against 2.58, and 4.87–5.82 against 5.15 —
and the search-cost intercept $\gamma_0$ stays near its true $-1.05$.

Under the normalization, $\gamma_0$ absorbs the misspecification and the damage
scales with how wrong the imposed value is. At a true spread of 2.58 the
estimate moves to roughly $-2.2$ to $-2.7$; at 5.15 it collapses to
**$-20$ to $-22$**. Since costs enter as $\exp(\gamma_0)$, that is a search cost
understated by many orders of magnitude — not a bias so much as a different
model.

## Result 2 — the counterfactual is destroyed, not merely biased

```{r}
knitr::kable(data.frame(
  `true sigma` = sprintf("%.2f", r$sig_true),
  `T` = r$T,
  `baseline clicks` = sprintf("%.2f", r$clicks_base),
  `true` = sprintf("%.2f", r$cf_true),
  `free` = sprintf("%.2f", r$cf_free),
  `normalized` = sprintf("%.2f", r$cf_norm),
  `error, free` = sprintf("%+.0f%%", 100 * r$cf_err_free),
  `error, normalized` = sprintf("%+.0f%%", 100 * r$cf_err_norm),
  check.names = FALSE), align = "r")
```

Errors are expressed as a share of the true predicted *change* in clicks, so
$-100\%$ means the model predicts no effect at all where a real one exists.

The free specification tracks the counterfactual within roughly 13% everywhere,
and within 5% in four of six cells. The normalized specification errs by
$-35\%$ at the smallest spread and by $-86\%$ to $-114\%$ once the true spread
exceeds two. At a true spread of 2.58 it is effectively predicting **that
halving search costs would do nothing**, when in fact clicks rise from 1.82 to
2.34.

This is the same failure @Yavorsky_2021 report, reproduced in a designed
instrument and, if anything, more severe. Their counterfactual was understated by
a factor of three to four; here the predicted effect is close to zero or
negative.

A related casualty is the position coefficient. Its true value is 0.12
throughout. Estimated freely it lands between 0.105 and 0.136 in every cell.
Under the normalization at a true spread of 5.15 it is estimated at **1.77 and
1.82** — an order of magnitude too large, because the position effect is being
asked to carry variation the spread should have explained.

## Result 3 — the panel is not what rescues the spread

Holding total tasks fixed, $T=8$ does not systematically beat $T=1$: the free
estimate of $\tilde\sigma$ is closer to the truth at $T=1$ in two of three
sweeps.

This is worth stating plainly because it sharpens the claim in
[`sections/identification.qmd`](../sections/identification.qmd). The
**exogenous shifter** is what identifies the post-search spread; the **panel** is
what identifies heterogeneity. These are separate arguments doing separate work,
and the model fitted here is homogeneous, so the panel has nothing to contribute.

The correct claim is therefore not "SIFT estimates $\tilde\sigma$ because it has
a panel" but "SIFT estimates $\tilde\sigma$ because position is randomized, and
separately supports individual-level parameters because it has a panel." Any
statement conflating the two should be corrected.

## Caveats

- One dataset per cell. The search-cost block carries substantial finite-sample
  bias (see `R/DIAGNOSTICS.md`), so individual cells should not be over-read.
  The pattern across cells is the result, not any single entry.
- Homogeneous fits throughout, which is why the panel arm is uninformative here.
  A heterogeneous version would test something different and is worth running.
- The normalization is imposed only in the reservation value. A researcher who
  additionally normalized $\sigma_\xi$ would produce different, probably larger,
  distortions.
