---
title: "Two instruments, one population"
subtitle: "Does conjoint mismeasure preference, or measure it well and fail downstream?"
format: html
bibliography: ../references.bib
execute: {echo: false, warning: false}
---

```{r}
#| include: false
res <- function(n) {
  p <- file.path("out", paste0(n, ".rds"))
  if (!file.exists(p)) p <- file.path("comparisons", "out", paste0(n, ".rds"))
  readRDS(p)
}
r <- res("instruments")
```

## The question

It is easy to say that conjoint "ignores search" and leave the implication that
its part-worths are therefore wrong. That claim needs testing, because there is a
competing account under which conjoint is doing nothing wrong at all.

A conjoint task shows every attribute of every alternative. The respondent sees
the post-click attributes and weighs them. Search governs what a shopper *learns*,
not what they *like*, so a full-information task may recover exactly the
preferences that operate in the market. Under that account conjoint measures
correctly and fails only when its estimates are carried into a world where
consideration is incomplete.

Two hypotheses, then:

- **A — conjoint mismeasures preference.**
- **B — conjoint measures preference well and fails downstream.**

## Design

One population with fixed preferences $(\boldsymbol\beta, \boldsymbol\kappa)$ and
fixed search costs. Both instruments are fielded on it.

| arm | task |
|---|---|
| SIFT | respondents search, inspect, and choose; fit the search model |
| conjoint | the same attribute profiles, all shown at once, no search; fit a conditional logit |

Then two separate questions. **Measurement:** does each instrument recover the
preference ratios? **Prediction:** does each reproduce market behaviour, where
behaviour is the search-then-choose process the population actually follows?

Comparison is on ratios rather than levels, since logit and probit scales differ.
Code: [`exp_instruments.R`](exp_instruments.R).

## Result 1 — conjoint measures preference *better* than SIFT

```{r}
m <- r$meas
knitr::kable(data.frame(
  ratio = m$ratio, truth = sprintf("%.2f", m$truth),
  SIFT = sprintf("%.3f", m$sift), conjoint = sprintf("%.3f", m$conjoint),
  `SIFT error` = sprintf("%+.3f", m$sift_err),
  `conjoint error` = sprintf("%+.3f", m$conjoint_err),
  check.names = FALSE), align = "r")
```

Hypothesis A is rejected, and firmly. The conjoint survey recovers every ratio to
within 0.03, including both post-click attributes. It is not attenuating anything.

SIFT recovers the listing-page ratio equally well but errs by $+0.44$ and
$-0.27$ on the two product-page ratios. The direction is unsurprising on
reflection: a conjoint task reveals the post-click attributes of every
alternative in every task, whereas SIFT reveals them only for alternatives the
respondent chose to open — roughly one in six here. Conjoint simply has far more
information about $\boldsymbol\kappa$ per task, and $\boldsymbol\kappa$ inherits
the finite-sample bias documented in `R/DIAGNOSTICS.md`.

**This is a cost of the instrument and should be reported as one.** SIFT is not a
better instrument for measuring preferences. On this evidence it is a somewhat
worse one.

## Result 2 — conjoint cannot represent the question that matters

```{r}
p <- r$pred
knitr::kable(data.frame(
  `listing position` = p$position,
  truth = sprintf("%.3f", p$truth),
  SIFT = sprintf("%.3f", p$sift),
  conjoint = sprintf("%.3f", p$conjoint),
  check.names = FALSE), align = "r")
```

Purchase share falls steadily down the listing page, from 0.178 at the top to
0.119 at the bottom — a ratio of **1.50**. SIFT reproduces this from its
estimates, 0.185 to 0.114, a ratio of **1.61**.

Conjoint predicts 0.155 at the top and 0.154 at the bottom. A ratio of
**1.00**.

This is not a bias that a larger sample would shrink. Conjoint has no position
parameter, because a conjoint task has no position: every alternative is equally
and costlessly visible. The model therefore *cannot* express the proposition that
moving an alternative up the page changes anything, and it predicts — with
complete internal consistency — that promoting a product from the bottom of the
listing to the top would have no effect whatsoever on its share.

## What this settles

**Hypothesis B, not A.** The failure is structural rather than statistical, and
it sits downstream of measurement.

Share is the product of the probability an alternative is considered and the
probability it is chosen once considered. Conjoint estimates the second term
accurately — better than SIFT does — and sets the first to one. Every question
about *what gets seen* therefore falls outside what the instrument can represent:
placement, ranking, merchandising, listing-page design, and the distinction
between an alternative that was rejected and one that was never opened.

The honest summary of the comparison is that the two instruments are good at
different things. Conjoint measures preference more precisely. SIFT measures
preference less precisely and adds a consideration stage that conjoint cannot
represent at any sample size. Which is preferable depends entirely on whether the
question is about preference or about share.

## Caveats

- One dataset per arm. The search-cost block carries substantial finite-sample
  bias, so the SIFT measurement errors here should not be read as a precise
  quantification of the penalty; replication would sharpen it.
- The conjoint respondent is modelled as fully compensatory, attending to every
  attribute. Real respondents screen, and non-compensatory processing would
  attenuate $\boldsymbol\kappa$ in the conjoint arm. This experiment therefore
  gives conjoint its **best case**, which is the right way to test hypothesis A.
- Both arms use the same attribute profiles and sample size, so the comparison
  holds the design fixed and varies only the task.
