---
title: "Conjoint vs SIFT"
subtitle: "What a conjoint analyst gets wrong when search is costly"
format: html
bibliography: ../references.bib
execute: {echo: false, warning: false}
---

```{r}
#| include: false
# works whether Quarto runs from the project root or from comparisons/
res <- function(n) {
  p <- file.path("out", paste0(n, ".rds"))
  if (!file.exists(p)) p <- file.path("comparisons", "out", paste0(n, ".rds"))
  readRDS(p)
}
```

## What this study is, and is not

This study fits a **full-information choice model to search-generated data** —
which is what one does when modelling field clickstream, and what a conjoint
share simulator implicitly assumes when it treats every alternative as
considered.

It is *not* a simulation of a conjoint survey. A conjoint survey constructs a
full-information task, so the respondent genuinely does see and weigh the
post-click attributes, and the part-worths it recovers need not be attenuated at
all. The distinction matters enough that it has its own study:
[two instruments, one population](compare-instruments.qmd) fields both
instruments on the same simulated respondents and separates measurement from
prediction.

Read the result below as a statement about *carrying* full-information
part-worths into a market where consumers search, not about what a conjoint
questionnaire measures.

## The question

SIFT nests conjoint. As the search cost goes to zero every alternative is
inspected, choice is made over fully realized utilities, and the model collapses
to a standard choice-based conjoint model. So the interesting question is not
whether the two differ in principle but **what a conjoint analyst gets wrong, by
how much, and in which direction**, when search costs are not zero.

## Design

Sweep the search-cost intercept from nearly free to expensive. At each level
simulate SIFT data (250 respondents × 8 tasks, $J=6$) and fit the model a
conjoint analyst would actually fit: a conditional logit over the **full** choice
set, using every attribute, with no notion of search.

Comparison is on **ratios** of coefficients rather than levels, because logit
and probit scales differ. Ratios are scale-free and are what a conjoint analyst
interprets — relative importance, willingness-to-pay.

The hypothesis worth testing is directional. Attributes visible only *after* a
click should be attenuated **more** than attributes visible on the landing page,
because conjoint assumes everyone saw everything, while in fact the post-click
attributes only influenced the choices of respondents who actually looked.

Code: [`exp_conjoint.R`](exp_conjoint.R).

```{r}
r <- res("conjoint")
knitr::kable(
  data.frame(
    `mean search cost` = r$res$cost,
    `mean clicks`      = round(r$res$mean_clicks, 2),
    `share of set seen`= sprintf("%.0f%%", 100 * r$res$frac_seen),
    `no purchase`      = sprintf("%.0f%%", 100 * r$res$frac_nobuy),
    `pre-click recovery`  = sprintf("%.2f", r$res$atten_pre),
    `post-click recovery` = sprintf("%.2f", r$res$atten_post),
    check.names = FALSE),
  align = "r")
```

The last two columns are the estimated coefficient ratio divided by its true
value. **1.00 means conjoint recovered the preference correctly; below 1 means
attenuated toward zero.**

## Result

The two columns behave completely differently.

**Pre-click attributes are recovered fine at every search cost.** The landing-page
ratio sits between 0.95 and 1.04 across the entire sweep, including where
respondents open only 2% of the choice set. Conjoint has no trouble with
attributes that respondents could see without clicking.

**Post-click attributes collapse monotonically.** Recovery falls from 0.93 when
search is nearly free, through 0.59, to 0.14 when clicking is expensive, and
turns slightly negative at the extreme. At a mean of 1.17 clicks out of 6 — an
inspection rate close to what @Yavorsky_2021 observes in the field, where mean
searches are 1.1 — **conjoint attenuates post-click part-worths by 64%**.

This is the paper's central claim made concrete. Conjoint is not uniformly
biased by search frictions; it is biased *selectively*, on exactly the
attributes that require inspection. An analyst running conjoint on a category
with real search costs will conclude that experience attributes barely matter,
when in fact respondents simply never saw them.

## Caveats

- One draw per cost level, so the numbers carry simulation noise; the monotone
  pattern is the result, not any individual entry.
- The conjoint model here is a homogeneous conditional logit. Random
  coefficients would absorb some — not all — of the attenuation, since the
  problem is a misspecified consideration set rather than unmodelled taste
  variation.
- The zero-search-cost limit is approached but not attained; at the cheapest
  level respondents still open only 45% of the set.
