---
title: "Ursu (2018) position analysis vs SIFT"
subtitle: "Can we test where position acts, instead of assuming it?"
format: html
bibliography: ../references.bib
execute: {echo: false, warning: false}
---

```{r}
#| include: false
# works whether Quarto runs from the project root or from comparisons/
res <- function(n) {
  p <- file.path("out", paste0(n, ".rds"))
  if (!file.exists(p)) p <- file.path("comparisons", "out", paste0(n, ".rds"))
  readRDS(p)
}
```

## The question

Every position-based search-cost shifter rests on one assumption: position
affects the **cost** of inspecting an alternative and nothing else. If position
instead shifted pre-search utility, it would be an attribute, not a cost
shifter, and the search-cost estimates built on it would be meaningless.

@Ursu_2018 §4.3 takes this seriously rather than asserting it, enumerating the
ways position could enter the model and ruling them out observationally. She has
to work hard because position on her platform is *assigned*. SIFT randomizes
position within every task, so the same tests become experimental contrasts.

The question here: **do the tests actually discriminate?**

## Design

Three data-generating processes, identical except for where position acts.

| arm | position … |
|---|---|
| `cost` | raises the search cost (the maintained assumption) |
| `utility` | lowers pre-search utility |
| `variance` | widens the post-search spread |

Two statistics, both computed from raw data with no model fitted:

- **T1** — among *clicked* alternatives, does position predict purchase?
- **T2** — is the first click near the top of the page?

Code: [`exp_position.R`](exp_position.R).

```{r}
r <- res("position")$res
knitr::kable(data.frame(
  `position acts on` = r$dgp,
  `mean clicks`      = round(r$mean_clicks, 2),
  `T1 slope`         = sprintf("%+.3f", r$t1_slope),
  `T1 95% CI`        = sprintf("[%+.2f, %+.2f]", r$t1_lo, r$t1_hi),
  `T2 first click`   = sprintf("%.2f", r$t2_first_pos),
  `T2 vs uniform`    = sprintf("%.1f", r$t2_uniform),
  check.names = FALSE), align = "r")
```

## Result

The tests discriminate — but **T1's sign is the opposite of the natural
intuition**, and getting it wrong would invert the conclusion.

The tempting reasoning is: if position acts only on costs, it should not bear on
what gets bought once someone has looked, so a *flat* T1 indicates
position-in-cost. That is wrong, because of selection. When position raises the
search cost, an alternative in a bad position must have unusually high
pre-search utility to be worth opening at all. Conditional on having been
clicked, bad-position alternatives are therefore *better* than good-position
ones, and T1 comes out strongly **positive** ($+0.39$, $p\approx5\times10^{-15}$).

When position enters pre-search utility instead, the direct effect and the
selection effect roughly cancel, and T1 is approximately **flat** ($-0.05$,
$p=0.06$) — the signature that the naive reading would have attributed to costs.

T2 separates the third case cleanly. A wider post-search spread *raises* the
reservation value, so the worst-placed alternatives get opened first: mean
first-click position 4.96 against a uniform benchmark of 3.5, versus 2.21 and
2.57 in the other two arms.

### The decision rule

| T2 | T1 | position acts on |
|---|---|---|
| near top | strongly positive | **search cost** |
| near top | flat | pre-search utility |
| near bottom | — | post-search spread |

Neither statistic identifies the case alone. Read jointly, they do.

## What this does and does not establish

It establishes what these two tests do **in this model**, and that SIFT's
randomization makes them available without any modelling assumption. That is
worth reporting as a design check in every SIFT study.

It does **not** establish what @Ursu_2018 concludes from conversion-on-rank in
her data. Her environment differs in ways that plausibly matter — position is
platform-assigned rather than randomized, and she observes one click per
impression — so the selection channel driving our T1 result may operate
differently there. The characterization of her argument in
[`sections/identification.qmd`](../sections/identification.qmd) carries a TODO
to check §4.3 directly before we attribute any sign pattern to her.
