---
title: "Estimating a Sequential Search Model"
subtitle: "A first course, from Weitzman's rule to a working estimator"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme:
      light: cosmo
      dark: darkly
bibliography: ../references.bib
execute:
  warning: false
  message: false
---

## Who this is for

You know maximum likelihood, discrete choice, and simulation-based estimation.
You have never estimated a search model. By the end you will have written one
from scratch, understood why the obvious approach fails, and seen where each
extension in the literature comes from.

Everything runs in base R. No packages.

## 1. Why a choice model is not enough

In a standard discrete choice model, a consumer sees $J$ alternatives, knows the
utility of each, and picks the best. The likelihood is
$\Pr(u_{iy} \ge u_{ij}\ \forall j)$, and with the right distributional assumption
it has a closed form.

Now suppose the consumer does not know the utilities. She sees something about
each alternative, and must pay a cost to learn the rest. She inspects a few, then
buys one of the ones she inspected. You observe **which alternatives she
inspected, in what order, how many, and which she bought.**

A choice model cannot use most of that. It has no notion of inspection, so it
cannot explain why she opened three alternatives rather than one, why she opened
*those* three, or why she stopped when she did. Worse, it will attribute her
purchase to a comparison she never made — against alternatives she never looked
at.

The search model's job is to explain all four margins at once: **which, in what
order, how many, and which bought.**

## 2. Pandora's boxes

::: {.callout-note}
## The problem @Weitzman_1979 solved

There are $J$ closed boxes. Box $j$ contains a prize $u_j$ drawn from a known
distribution $F_j$. Opening it costs $c_j$. You may open boxes in any order,
stopping whenever you like, and you keep the best prize you have seen. What is
the optimal policy?
:::

This looks like it should require dynamic programming over the set of opened
boxes — an exponentially large state space. Weitzman's result is that it does
not.

### The reservation value

For each box, define $z_j$ as the prize level that would make you exactly
indifferent about opening it:

$$
\mathbb{E}\left[\max(u_j - z_j,\ 0)\right] \;=\; c_j .
$$

Read the left side as *the expected gain from opening box $j$ when you already
hold $z_j$ in hand*. If you hold more than $z_j$, opening is not worth it. If you
hold less, it is.

::: {.callout-important}
## The result that makes everything tractable

$z_j$ depends **only on box $j$'s own distribution and its own cost**. It does
not depend on what else is available, on how many boxes there are, or on what you
have already found.

That is what collapses the dynamic program. Each box gets a single number,
computed in isolation, once and for all.
:::

### Working out $z$ for the normal case

Let $u_j \sim N(\delta_j, \sigma^2)$. Substituting $t = (u-\delta_j)/\sigma$ and
$\zeta_j = (z_j - \delta_j)/\sigma$:

$$
\mathbb{E}\left[\max(u_j - z_j, 0)\right]
= \sigma \int_{\zeta_j}^{\infty} (t - \zeta_j)\,\phi(t)\,dt
= \sigma\Big[\underbrace{\phi(\zeta_j) - \zeta_j\big(1 - \Phi(\zeta_j)\big)}_{\displaystyle g(\zeta_j)}\Big].
$$

So the reservation value is

$$
z_j = \delta_j + \zeta_j\,\sigma, \qquad \text{where } g(\zeta_j) = c_j/\sigma .
$$

$g$ is strictly decreasing, so it inverts. Cheap search means a large $\zeta$ and
a reservation value well above the mean; expensive search pulls $\zeta$ down.

```{r}
g     <- function(z) dnorm(z) - z * pnorm(z, lower.tail = FALSE)
g_inv <- function(k) uniroot(function(z) g(z) - k, c(-10, 10))$root

curve(g, -2, 3, lwd = 2, xlab = expression(zeta), ylab = expression(g(zeta)),
      main = "g is strictly decreasing, so it inverts")
abline(h = 0.2, lty = 2); abline(v = g_inv(0.2), lty = 2)
```

```{r}
c(g_of_0 = g(0), zeta_when_cost_is_0.2 = g_inv(0.2))
```

## 3. Weitzman's three rules

With reservation values in hand, optimal behaviour is:

1. **Selection.** Inspect in decreasing order of $z_j$.
2. **Stopping.** Stop when the best prize in hand exceeds the highest $z$ among
   the unopened boxes.
3. **Choice.** Take the best prize you have opened.

Everything we do from here is the statistics of these three rules.

## 4. The model we will estimate

Consumer $i$ faces $J$ alternatives.

$$
\underbrace{\delta_{ij} = \mathbf{x}_{ij}'\boldsymbol\beta + \eta_{ij}}_{\text{known before searching}},
\qquad
\underbrace{u_{ij} = \delta_{ij} + \varepsilon_{ij}}_{\text{learned by searching}}
$$

with $\eta_{ij}\sim N(0,1)$ and $\varepsilon_{ij}\sim N(0,1)$, and a **constant,
known-form search cost** $c$. We start with both variances fixed at one; §9.2
takes up estimating the second.

::: {.callout-warning}
## Why we need $\eta$

Without $\eta$, $\delta_{ij}$ is a deterministic function of $\mathbf{x}_{ij}$,
so the search order is deterministic too. Any observed order that does not match
the ordering of $\mathbf{x}_{ij}'\boldsymbol\beta$ would have probability exactly
zero, and the likelihood would be degenerate. **Reservation values must carry a
stochastic component** or the model cannot rationalize observed search orders
[@Ursu_Seiler_Honka_2024].
:::

Note a useful simplification: with $c$ constant and $\sigma=1$, every alternative
has the *same* $\zeta$. So $z_{ij} = \delta_{ij} + \zeta$ is a constant shift, and

$$
\textbf{ordering by } z \;=\; \textbf{ordering by } \delta .
$$

Search order reveals the ordering of pre-search utilities. The *number* of
searches is what speaks to $c$. Keep that distinction in mind — it is the whole
of the identification argument in §8.

## 5. Ten consumers

```{r}
sim <- function(N, J, beta, cost, seed = 1) {
  set.seed(seed)
  d <- data.frame(id = rep(1:N, each = J), alt = rep(1:J, N),
                  x1 = rnorm(N * J), x2 = rnorm(N * J))
  d$eta   <- rnorm(N * J)
  d$eps   <- rnorm(N * J)
  d$delta <- d$x1 * beta[1] + d$x2 * beta[2] + d$eta
  d$u     <- d$delta + d$eps
  zeta    <- g_inv(cost)
  d$z     <- d$delta + zeta

  searched <- vector("list", N); choice <- integer(N)
  for (i in 1:N) {
    r   <- d[d$id == i, ]
    ord <- order(r$z, decreasing = TRUE)   # rule 1: search in decreasing z
    K <- 1L; best <- r$u[ord[1]]
    while (K < J && best < r$z[ord[K + 1L]]) {   # rule 2: stop when best >= next z
      K <- K + 1L; best <- max(best, r$u[ord[K]])
    }
    S <- ord[1:K]
    searched[[i]] <- S
    choice[i] <- S[which.max(r$u[S])]      # rule 3: buy the best inspected
  }
  list(d = d, S = searched, y = choice, zeta = zeta,
       N = N, J = J, beta = beta, cost = cost)
}

dat <- sim(N = 10, J = 3, beta = c(1.0, -0.5), cost = 0.2, seed = 7)
```

What an analyst actually observes — the covariates, the ordered inspection
sequence, and the purchase:

```{r}
obs <- data.frame(
  consumer = 1:dat$N,
  searched = sapply(dat$S, paste, collapse = " > "),
  n_searched = sapply(dat$S, length),
  bought = dat$y)
knitr::kable(obs)
```

Look at consumer 1. She opened alternative 1 and stopped. That single row says
three things: $z_1$ was the highest of the three; whatever she found was good
enough to beat both remaining reservation values; and she bought it. Two
inequalities and a choice, from one consumer.

Consumer 3 opened all three. That says her first two draws were *disappointing*
— each time, the best in hand stayed below the next reservation value.

::: {.callout-tip}
## The data you never see

$\eta$ and $\varepsilon$ are hidden, and $\varepsilon$ is only *realized* for
alternatives she opened. Nothing in the data reveals the match value of an
alternative she never inspected. That is the whole estimation problem.
:::

## 6. The likelihood

For consumer $i$ with ordered search set $S = (s_1,\dots,s_K)$, unsearched set
$\bar S$, and purchase $y$, the four conditions are:

| condition | inequality |
|---|---|
| **selection** | $z_{s_1} \ge z_{s_2} \ge \dots \ge z_{s_K} \ge \max_{l \in \bar S} z_l$ |
| **continuation** | $\max_{h<k} u_{s_h} < z_{s_k}$ for $k = 2,\dots,K$ |
| **stopping** | $\max_{h \le K} u_{s_h} \ge \max_{l \in \bar S} z_l$ |
| **choice** | $u_y \ge u_{s_h}$ for all $h \le K$ |

and the contribution to the likelihood is the probability that all hold:

$$
\Pr_i(\boldsymbol\beta, c) = \int \mathbb{1}\{\text{all four}\}\; dF(\boldsymbol\eta)\, dF(\boldsymbol\varepsilon).
$$

**This integral has no closed form.** It is a probability over a region of
$\mathbb{R}^{2J}$ carved out by inequalities that mix $\eta$ (through both $z$ and
$u$) and $\varepsilon$ (through $u$ only). Unlike probit, the region is not an
orthant in any convenient basis.

So we simulate it.

## 7. First attempt: the crude frequency simulator

The obvious estimator. Draw $\boldsymbol\eta$ and $\boldsymbol\varepsilon$, check
whether all four conditions hold, and count.

```{r}
loglik_cfs <- function(par, dat, R = 2000, seed = 99) {
  set.seed(seed)
  b <- par[1:2]; cost <- exp(par[3])
  zeta <- g_inv(cost); N <- dat$N; J <- dat$J
  ll <- 0
  for (i in 1:N) {
    r  <- dat$d[dat$d$id == i, ]
    xb <- r$x1 * b[1] + r$x2 * b[2]
    S  <- dat$S[[i]]; K <- length(S); U <- setdiff(1:J, S); y <- dat$y[i]
    hit <- 0
    for (rep in 1:R) {
      delta <- xb + rnorm(J)
      u     <- delta + rnorm(J)
      z     <- delta + zeta
      ok <- TRUE
      if (K >= 2) for (k in 1:(K - 1)) if (z[S[k]] < z[S[k + 1]]) { ok <- FALSE; break }
      if (ok && length(U)) if (z[S[K]] < max(z[U])) ok <- FALSE
      if (ok && K >= 2) for (k in 2:K)
        if (max(u[S[1:(k - 1)]]) >= z[S[k]]) { ok <- FALSE; break }
      if (ok && length(U)) if (max(u[S]) < max(z[U])) ok <- FALSE
      if (ok) if (u[y] < max(u[S])) ok <- FALSE
      hit <- hit + ok
    }
    ll <- ll + log(max(hit / R, 1e-10))
  }
  ll
}

loglik_cfs(c(1.0, -0.5, log(0.2)), dat)
```

It works, and it is unbiased for each probability. **It is also unusable for
optimization**, and the reason is worth seeing rather than being told.

Hold the draws fixed and vary $\beta_1$ on a fine grid, for a single consumer:

```{r}
#| fig-height: 4
one <- dat; one$N <- 1
one$d <- dat$d[dat$d$id == 1, ]; one$S <- dat$S[1]; one$y <- dat$y[1]

grid <- seq(0.6, 1.4, by = 0.01)
cf   <- sapply(grid, function(b) loglik_cfs(c(b, -0.5, log(0.2)), one, R = 200))

plot(grid, cf, type = "s", lwd = 2, xlab = expression(beta[1]),
     ylab = "log-likelihood", main = "Crude frequency: a staircase")
```

```{r}
c(distinct_values = length(unique(round(cf, 8))),
  grid_points     = length(grid),
  longest_flat_run = max(rle(round(cf, 8))$lengths))
```

The estimator is a **step function**. Its derivative is zero almost everywhere
and undefined at the jumps, so gradient-based optimizers have nothing to follow.
The indicator either fires or it does not; small parameter changes move no draw
across a boundary until suddenly one does.

## 8. The fix: smooth the indicator

Replace the hard indicator with a smooth approximation. Write each condition as a
margin $\nu \ge 0$ that should be non-negative, and use a logit kernel
[@Yavorsky_2021]:

$$
\tilde{\mathbb{1}} \;=\; \Big(1 + \textstyle\sum_m e^{-\lambda \nu_m}\Big)^{-1}.
$$

When every margin is comfortably positive this is near 1; when any is negative it
is near 0; and in between it varies **smoothly** with the parameters. The scale
$\lambda$ controls the trade-off: larger is a better approximation to the
indicator but a less smooth objective.

```{r}
loglik_smooth <- function(par, dat, R = 500, lambda = 8, seed = 99) {
  set.seed(seed)
  b <- par[1:2]; cost <- exp(par[3])
  zeta <- tryCatch(g_inv(cost), error = function(e) NA)
  if (is.na(zeta)) return(-1e10)
  N <- dat$N; J <- dat$J
  ll <- 0
  for (i in 1:N) {
    r  <- dat$d[dat$d$id == i, ]
    xb <- r$x1 * b[1] + r$x2 * b[2]
    S  <- dat$S[[i]]; K <- length(S); U <- setdiff(1:J, S); y <- dat$y[i]
    acc <- numeric(R)
    for (rep in 1:R) {
      delta <- xb + rnorm(J)
      u     <- delta + rnorm(J)
      z     <- delta + zeta
      pen <- 0
      # selection
      if (K >= 2) for (k in 1:(K - 1)) pen <- pen + exp(-lambda * (z[S[k]] - z[S[k + 1]]))
      if (length(U)) for (l in U)      pen <- pen + exp(-lambda * (z[S[K]] - z[l]))
      # continuation
      if (K >= 2) for (k in 2:K)       pen <- pen + exp(-lambda * (z[S[k]] - max(u[S[1:(k - 1)]])))
      # stopping
      if (length(U))                   pen <- pen + exp(-lambda * (max(u[S]) - max(z[U])))
      # choice
      for (h in S) if (h != y)         pen <- pen + exp(-lambda * (u[y] - u[h]))
      acc[rep] <- 1 / (1 + pen)
    }
    ll <- ll + log(max(mean(acc), 1e-300))
  }
  ll
}
```

The same slice, now smooth:

```{r}
#| fig-height: 4
sm <- sapply(grid, function(b) loglik_smooth(c(b, -0.5, log(0.2)), one, R = 200))
plot(grid, sm, type = "l", lwd = 2, col = "steelblue", xlab = expression(beta[1]),
     ylab = "log-likelihood", main = "Smoothed: differentiable")
abline(v = 1.0, lty = 2)
```

```{r}
c(distinct_values = length(unique(round(sm, 8))),
  longest_flat_run = max(rle(round(sm, 8))$lengths))
```

Every grid point gives a distinct value. There is now a gradient to follow.

::: {.callout-warning}
## What smoothing costs

The smoothed estimator is **biased** for any finite $\lambda$: it is only
asymptotically unbiased as $\lambda \to \infty$. But large $\lambda$ restores the
staircase. You are choosing a point on a bias–smoothness trade-off, and
$\lambda$ is a tuning constant you must justify. §9.5 describes estimators that
avoid it.
:::

## 9. Does it recover the truth?

Ten consumers is too few to estimate three parameters. Six hundred is enough to
see recovery.

```{r}
#| cache: true
big <- sim(N = 600, J = 3, beta = c(1.0, -0.5), cost = 0.2, seed = 21)
prop.table(table(sapply(big$S, length)))
```

```{r}
#| cache: true
fit <- optim(c(0, 0, log(0.5)),
             function(p) -loglik_smooth(p, big, R = 300, lambda = 8),
             method = "BFGS", control = list(maxit = 200))

data.frame(parameter = c("beta1", "beta2", "search cost"),
           truth     = c(1.0, -0.5, 0.2),
           estimate  = c(fit$par[1], fit$par[2], exp(fit$par[3])))
```

Both preference parameters and the search cost are recovered. Note that we
estimated $\log c$ rather than $c$, so that the optimizer cannot propose a
negative cost.

## 10. What identifies what

Worth pausing on, because it is not obvious and it drives every design decision
in applied work.

**Preferences ($\boldsymbol\beta$) are identified twice over.** They govern the
purchase among inspected alternatives — the same margin a choice model uses — and
they govern the *order* of inspection, since $z$ and $\delta$ order identically.
Two independent sources of information.

**The search cost is identified by how much people search, and by almost nothing
else.** Raise $c$ and reservation values fall, so consumers stop sooner. The
distribution of $K$ across consumers is the moment doing the work.

This asymmetry has consequences you will meet immediately in practice:

- $\boldsymbol\beta$ is estimated precisely; $c$ is not.
- If most consumers inspect exactly one alternative, there is very little
  variation in $K$ and the search cost is weakly identified.
- Anything that shifts search costs *exogenously and across alternatives* is
  extremely valuable, because it produces variation in $K$ that is not
  confounded with preference. This is why @Ursu_2018 makes so much of randomized
  rankings, and why @Yavorsky_2021 builds an argument around driving distance.

## 11. Extensions

Each of these is one step from what you have already written.

### 11.1 Search costs that depend on covariates

Replace the constant with $c_{ij} = \exp(\mathbf{w}_{ij}'\boldsymbol\gamma)$.
Positivity comes free from the exponential. In the code, `zeta` stops being a
scalar and becomes a vector — call `g_inv` per alternative.

The consequence is more interesting than the code change. With a constant cost,
$\zeta$ was common across alternatives, so search order was determined entirely
by $\delta$. Now alternatives with different costs have different $\zeta$, and
**search order is no longer the same as preference order**. An alternative may be
inspected early because it is cheap to inspect rather than because it is
attractive.

That is a loss of a clean identification argument, and simultaneously the source
of the best identification argument available: if $\mathbf{w}$ contains something
randomized, it moves search order without moving preference.

### 11.2 Estimating the match-value standard deviation

We fixed $\sigma_\varepsilon = 1$. Almost all of the literature does the same.
Why?

Look again at $g(\zeta) = c/\sigma$ and $z = \delta + \zeta\sigma$. The search
cost and the match-value spread enter through the *same channel*, and they push
in opposite directions: a large $\sigma$ makes searching attractive, a large $c$
makes it unattractive. Both change $K$.

They are separately identified by functional form, since $\sigma$ appears on both
sides of the $g$ equation — but functional form alone is weak. In practice, the
two are nearly collinear: @Yavorsky_2021 report a correlation of $0.95$ between
$\log\sigma$ and the search-cost intercept even with 6,511 consumers.

What breaks the tie is an *exogenous, alternative-specific* search-cost shifter.
With one in hand, $\boldsymbol\gamma$ is identified from how search responds to
$\mathbf{w}$, and $\sigma$ is then identified separately from the level of
search. @Yavorsky_2021 estimate $\sigma = 8.16$ where the literature had been
imposing 1, with large consequences for what search costs are worth in money.

The exercise to run: fix $\sigma$ at 1 in your estimation while simulating data
with $\sigma = 3$, and watch the search-cost estimate absorb the misspecification.

### 11.3 An outside option, and zero searches

We assumed the first search always happens. Real data include consumers who
inspect nothing.

Add an alternative $j=0$ with utility known from the outset and no search
required. The continuation rule at $k=1$ then acquires content — the best
reservation value must beat the outside option — and $K=0$ becomes an observed
outcome rather than something you must condition away.

This matters more than it looks. Conditioning on having searched selects the
sample on an outcome the model is trying to explain, and papers that do it must
assume the first search is free to avoid the resulting bias.

### 11.4 Heterogeneity

Let $\boldsymbol\beta_i$ and $\boldsymbol\gamma_i$ vary across consumers, with
$\boldsymbol\theta_i \sim N(\bar{\boldsymbol\theta}, \Sigma)$.

The obstacle is that identifying individual-level parameters requires observing
each consumer **more than once**. With one search spell per consumer, you can
estimate the distribution's parameters but not each consumer's position in it.
@Morozov_2021 prove nonparametric identification from panel search data, with the
caveat that the argument takes sessions per consumer to infinity — with a mean of
five they report hyperparameters only.

Two estimation routes. Simulated maximum likelihood integrates over the
heterogeneity distribution, multiplying cost by the number of mixing draws.
Hierarchical Bayes does not: each consumer sits at their own $\boldsymbol\theta_i$
and the mixing-draw multiplier disappears. For search models, **HB is usually the
cheaper route**, which is the reverse of the folklore.

### 11.5 Better simulators

Kernel smoothing is general and easy to write, which is why it dominated the
field for a decade. Its costs are the tuning constant $\lambda$, the bias for
finite $\lambda$, and high simulation variance.

Three alternatives, roughly in order of how much machinery they require:

- **Importance sampling** [@Ackerberg_2009] reuses draws across parameter values,
  so the likelihood becomes smooth in the parameters without a kernel.
- **GHK-type simulators** exploit the fact that, *conditional on the observed
  purchase*, the four conditions become a conjunction of linear inequalities —
  an orthant probability, which GHK evaluates far more accurately than
  accept–reject.
- **Reformulating what is random** [@Chung_2025] makes the search *cost*
  stochastic rather than the pre-search taste shock, which turns the search
  probabilities into recursive truncated draws and reports speedups of one to two
  orders of magnitude.

@Ursu_Seiler_Honka_2024 compare these systematically and is the right next thing
to read.

## 12. Exercises

1. Set $\beta = 0$ and re-simulate. What does the distribution of $K$ look like,
   and can you still recover $c$?
2. Drive $c \to 0$. Verify that everyone inspects everything and the model
   collapses to a standard choice model over fully realized utilities. This is
   the sense in which conjoint is the zero-search-cost special case.
3. Vary $\lambda$ over $\{2, 8, 30\}$ and plot the objective. Where does the
   staircase reappear?
4. Simulate with $\sigma_\varepsilon = 3$, estimate with it fixed at 1, and
   report the bias in $\hat c$. Then add a randomized cost shifter and estimate
   $\sigma$ freely.
5. Add an outside option and allow $K=0$. How much does the search-cost estimate
   move when the zero-search consumers are included rather than dropped?
